Automated Lensing Learner - I: An automated Strong Lensing Identification Pipeline

### 2017-03-23 ms.tex

%-------------------------------------------------%
\begin{abstract} 
Gravitational lensing offers a direct probe of the underlying mass
distribution of lensing systems, a window to the high redshift
universe, and a geometric probe of cosmological models.  The advent of
large scale surveys such as the Large Synoptic Sky Telescope and
Euclid has prompted a need for automatic and efficient identification
of strong lensing systems.  We present (1) {\em (ALL) Automated
  Lensing Learner}, a strong lensing identification pipeline that will
be publicly released as open source software, and (2) a publicly
available mock LSST dataset with strong galaxy-galaxy lenses.  In this
first application of the pipeline, we employ a fast feature extraction
method, Histogram of Oriented Gradients (HOG), to capture edge
patterns that are characteristic of strong gravitational arcs in
galaxy-galaxy lensing.  We use logistic regression to train a
supervised classifier model on the HOG of HST- and LSST-like images.
We use an area under the curve (AUC) of a Receiver Operating
Characteristic curve to assess model performance; $AUC=1.0$ is an
ideal trained classifier, and $AUC=0.5$ does not provide any
classification information.  Our best performing models on a training
set of 10,000 lens containing images and 10,000 non-lens containing
images exhibit an AUC of 0.975 for an HST-like sample.  However, for
one orbit of LSST, our model only reaches an AUC of 0.625.  For
10-year mock LSST observations, the AUC improves to 0.809.  Model
performance appears to continually improve with the size of the
training set.  Models trained on fewer images perform better in
absence of the lens galaxy, the removal of which can decrease the
false positive rate due to the blending of light between source images
and the lens galaxy.  However, with larger training data sets,
information from the lens galaxy improves model performance.  Our
results demonstrate efficient and effective methods for automated
strong lensing identification. With existing open source tools, this
example linear classifier both runs on a personal laptop and can
easily scale to large data sets on a computing cluster.
\end{abstract}
%-------------------------------------------------%

@@

% Gravitational lensing as a geometric test of cosmology.  Also
% probes underlying matter distribution in galaxy clusters, sensitive
% to things like recent accretion and environment.
Gravitational lensing occurs when intermediate fluctuations in the
matter density field deflect light from background sources
\citep[see][for a review]{kneibandnatarajan_11}.  Strong gravitational
lensing can manifest as visible giant arcs magnifying high redshift
galaxies \citep{lyndsandpetrosian_86,gladders_etal03}, multiply imaged
quasars \citep{walsh_etal79}, multiply imaged galaxies
\citep{sharon_etal05}, and arclets \citep{bezecourt_etal98}.  Lensing
signatures probe the underlying dark matter distribution of the lens
\citep{warrenanddye_03}, high redshift galaxy formation
\citep{allam_etal07}, and provide a geometric test of cosmology via
comparison of predicted arc abundances and observed abundances
\citep{kochanek_96,chae_03,linder_04} and time-delay signals from
multiply imaged quasars\citep{xli_etal12,suyu_etal14,suyu_etal16}.

% Application of constraining mass profiles - large numbers needed.
The application of strong gravitational lensing to constrain the mass
distribution of strong lenses, such as early-type galaxies (ETGs),
necessitates large samples of galaxy-galaxy lensing systems.
\citet{miraldaescudeandlehar_92} first suggested that massive
ellipticals would be likely frequent strong lensing sources in optical
surveys.  These systems exhibit a background source galaxy that the
lens galaxy deflects into a partial of full arc shaped Einstein ring.
The strong lensing signature directly probes the underlying matter.
The identification of such systems in upcoming surveys is the first
step in constraining the mass-to-light ratio for a large number of
objects in this mass range.

@@

% Transition to automated identification
Recent efforts have focused on the development of automated methods
with performance comparable to or better than humans. {\em SpaceWarps}
is a part of the Zooniverse Project \citep{marshall_etal16}, which
also includes {\em Galaxy Zoo}, the citizen science based image
classification of galaxy types \citep{lintott_etal08}.  Galaxy
classification is an early successful example of in which machine
learning algorithms successfully train models to classify astronomical
images with comparable performance to humans \citep{dieleman_etal15}.

% Early efforts
Earlier efforts on automated, or ``robot'', identification of strong
lensing systems have two distinct generalized steps.  The first
enhances and extracts characteristic features, and the second uses
some form of pattern recognition in the features to classify lens and
non-lens containing systems.  Among others, selected features might
include shape parameterization
\citep[see][e.g.]{,alard_06,kuboanddellantonio_08,xu_etal16,lee_17},
colors in multi-band imaging \citep{maturi_etal14,gavazzi_etal14},
light profiles \citep{braultandgavazzi_15}, and characteristics of
potential lens galaxies \citep{marshall_etal09}.  Pattern recognition
often incorporate cutoffs in selected parameter space distributions
\citep{lenzen_etal04,joseph_etal14,paraficz_etal16}.  A number of
these have been publicly distributed, with specific end applications.
For example, {\em ArcFinder} is one such code that finds arcs in
groups or cluster scale lens \citep{seidelandbartelmann_07}, and {\em
  RingFinder} an analogous tool in searching for multiply imaged
quasars \citep{gavazzi_etal14}.  Codes like these have been
complementary to human identification \citep{more_etal16}.

% Classical machine learning methods
The latter of these methods have transitioned to using ``machine
learning'' algorithms for the pattern recognition step.  In classical
machine learning, we can train a model to separate a dataset according
to some desired classification scheme.  However, we still need to
reduce the dimensionality of the problem with a feature extraction
method.  The dominant ``features'' of that determine image
classification are not a priori known, but determined by the
algorithm.

% Recent works, neural nets, and deep learning
Much more recent works have made use of a subset tool in machine
learning, neural networks, to classify images either from input image
parameters or directly in image space.  In \citet{estrada_etal07},
authors used input shape parameters to train neural nets to identify
arc candidates.  \citet{agnello_etal15} used neural networks trained
on data from multiband magnitudes, and \citet{bom_etal16} used
extracted morphological parameters.

% Convolutional and residual
The most advanced use of neural networks for strong lensing
classification operate directly in image space.
\citep{petrillo_etal17} used mock Kilo Degree Survey data to train
convolutional neural networks, with a training set size of six million
images. \citet{lanusse_etal17} used state-of-the-art residual learning
to also work directly in the image space with minimal image
pre-processing.  A major strength of the \citet{lanusse_etal17}
implementation is that residual learning has been found to be easier
to train in deep neural networks, and can outperform implementations
of convolutional neural networks.

%  Relevance of our work
While there has been a recent surge in the use of deep neural
networks applied to image classification problems in astronomy,
it is not always easy to scale these techniques to large data
sets, nor are the necessary computational resources and hardware,
such as GPUs, easily accessible to the
entire scientific community.  We present a supervised classification
pipeline built with open source tools, tested on mock HST and LSST
data, which will also be made publicly available.  Our goal is to
centralize an open source pipeline that is general enough for the
user to select or add appropriate image processing techniques to
augment the features of the lensed image, train and test data with
different machine learning techniques, and to select the sequence of
methods that provides optimal model performance.  Results are
reproducible on personal computers, as both the pipeline and the
data will be publicly distributed.

% Summary of paper 
This first paper serves as both a presentation of the mock data set
and as a thorough case study for our pipeline with one feature
extraction and classical machine learning method, respectively, the
{\em histogram of oriented gradients} (HOG) and {\em logistic
  regression} (LR).  HOG is a fast feature extraction procedure that
quantifies edges in images, commonly used to identify humans in
security software.  LR is a linear classifier, making its scalability
relatively straightforward with existing open source tools.  This
framework is easily extendable to other feature methods and machine
learning algorithms for supervised classification.

@@

There are 10,000 lens containing mock observations and 10,000
non-lens containing mock observations for running our parameter search. 
We keep a hold-out test set of 1,000 lens containing mock observations
and 1,000 non-lens containing mock observations on which we evaluate
the final trained model.

@@

To produce simulated lensed galaxies, we first model the mass of lens
galaxies as the Singular Isothermal Ellipsoid (SIE). This model is
both analytically tractable and is consistent with models of
individual lenses and lens statistics on the length scales relevant
for strong lensing
\citep[e.g.]{koopmans_etal06,gavazzi_etal07,dye_etal08,liandchen_09}.

The normalized convergence map of SIE model is defined as:
%-------------------------------------------------%
\begin{equation}\label{eqn:sie}
\kappa = \frac{\theta_{\rm E}}{2}\frac{1}{\sqrt{x_1^2/q+x_2^2 q}},
\end{equation}
%-------------------------------------------------%
where $\theta_{\rm E}$ is the Einstein Radius and $q$ is axis ratio.
The Einstein radius can be calculated from the redshift of the
lens, the redshift of the source, and the velocity dispersion of the
lens galaxy as follows,
%-------------------------------------------------%
\begin{equation}
\theta_{\rm E} = 4\pi\left(\frac{\sigma_v}{c}\right)\frac{D_{ls}}{D_{s}}, 
\end{equation}
%-------------------------------------------------%
Here, $c$ is the speed of light, $\sigma_v$ is the velocity dispersion
of the lens galaxy, $D_{ls}$ and $D_{s}$ are respectively the angular
diameter distances from the source plane to the lens plane and from
the source plane to the observer.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure*}[t]\label{fig:mockimages}
\begin{center}
\includegraphics[width=2\columnwidth]{figures/tiles/compilations/telescope_compilation.pdf}
\caption{Left to right show example mock HST, LSST 10 year, and LSST 1
  year images.  The top row corresponds to a lensing system with a
  very visible arc signature, and the bottom row to a lensing system
  that is less obvious.  Example mock HST images have
  $n_\text{pix}\times n_\text{pix}=300\times300$.  Example mock LSST
  images have $n_\text{pix}\times n_\text{pix}=45\times45$.  The
  resolution and noise of a ground based telescope is noticeably
  worse. Visual identification of giant arcs in the LSST images in the
  bottom row is very difficult.}
\end{center}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

To perform our analysis, we have used tools from {\em Scikit-learn}
\citep{pedregosa_etal12}.  We outline the identification pipeline in
Figure~\ref{fig:pipeline}, which is a general description of
supervised classification.  Supervised classification is a class of
machine learning where the labels of classification in the training
set are known.  In our case, the labels are lens and non-lens
containing images.

The first step of our pipeline consists of a feature extraction stage,
where our feature vector is a {\em histogram of oriented gradients}
\citep{dalalandtriggs_05} that quantifies edges in the image.  We
describe the method and parameter search in Section~\ref{sec:hog}.  We
then use {\em Logistic Regression} (LR), a machine learning algorithm
described in Section~\ref{sec:LR}, to train a classifier model on a
subset of our images.  LR requires a parameter search over the
regression coefficient, $C_{LogReg}$, which we explore in
Section~\ref{sec:regularization}.  We briefly comment that our initial
tests with {\em Support Vector Machine} (SVM) as an alternative
machine learning algorithm yielded negligible performance improvement,
and significantly increased computation cost.  This indicated that the
features of lens and non-lens images are relatively well separated by
hyperplanes in feature space.  For these reasons, we do not include
SVM in our final analysis and comparisons and continue all discussions
with a linear classifier.

@@

The second step of our analysis is to test our trained model on an
independent subset of the images to assess the model performance.
Here, we evaluate the model on each test image, predicting a
likelihood (``score'') between 0 and 1 that this is a lens containing
system.  This ``holdout set'' is not used in any of our parameter
searches to keep our test metric independent of tuning.

@@

%-------------------------------------------------%
\subsubsection{Feature Extractor: Histogram Oriented Gradients}\label{sec:hog}
%-------------------------------------------------%
% HOG intro - 
Originally created for human detection in computer vision, histogram
oriented gradients (HOG) is a feature extraction method that computes
centered horizontal and vertical gradients. HOG is relatively robust
to noise in the image, and is a fairly fast transform that describes
edges.  Details can be found in \citet{dalalandtriggs_05}, but we
describe the procedure here.  The end result of HOG is a one
dimensional histogram computed as follows.

@@

% Initial binning
The orientation gradient of all pixels within each cell are binned
into the quantized orientations, providing a net gradient description
within that cell.  As an example, for $N_{orient}=3$, our bins are
centered at $\theta=0, 2\pi/3, 4\pi/3$ in radians.  If a cell has no
gradient in the vertical direction, and only a gradient across the
horizontal direction, say in the positive x-direction with
$\theta=\pi/2$, it will contribute 75\% of its magnitude to the
$\theta=2\pi/3$ bin, and 25\% of its magnitude to the $\theta=0$ bin.
The bins in all cells are then concatenated to make a larger feature
vector that is $N_{orient}$\times$N_{cells}$.

% Normalization
The last step is a normalization procedure to control for illumination
effects.  Here, the sub-histograms of each cell within the same block
are normalized with respect to one another before the transformation
returns the final feature vector.  The division of the image and the
quantization of orientations provide a total of 3 parameters in HOG:
$N_{orient}$, cells-per-block, and pixels-per-cell.  We discuss how we
select parameters using cross-validation in
Section~\ref{sec:gridsearch}

@@

Next, the cells per block parameter determines the normalization of
each cell with respect to the neighboring cell.  In general, this will
downweight arc-like edges in cells that neighbor very bright cells,
such as cells that cover the central lens galaxy.  We therefore vary
the cells per block parameter between (2, 2) and (4, 4) for the
LSST-like images and between (1, 1) and (4, 4) for the HST-like
images.

@@

Finally, the resolution of the overall image will also limit the
additional information that an increase in $N_\text{orient}$ will
provide.  The best case number of orientations for each dataset is
$N_\text{orient,HST}=9$, $N_\text{orient,LSST10}=3$, \todo{and
  $N_\text{orient,LSST1}=$.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure*}[t]\label{fig:hogimages}
\begin{center}
\includegraphics[width=2\columnwidth]{figures/tiles/compilations/hog_telescope_compilation.pdf}
\caption{Left to right show example image transforms of mock images
  from Figure~\ref{fig:mockimages} with a visualized histogram of
  oriented gradients.  The image transform picks up edge features,
  with arc features showing up as edges across radial orientations.
  \wording{Each of the oriented gradients within a cell is
    color-coded by magnitude, and represented as a line in the
    direction perpendicular to that gradient.}  The actual extracted
  features fill a one-dimensional feature vector comprised of the
  magnitudes of each of the oriented edges within the visualized
  cells.}
\end{center}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%            
%% \caption{Top row: Visualized HOG feature vectors of the sample
%%   lensed HST-like image from Figure~\ref{fig:mockimages}.  Left:
%%   The corresponding HOG visualization with N$_\text{orient}=5$,
%%   (32, 32) pixels per cell, and (1, 1) cells per block.  Right: The
%%   HOG visualization of the same mock image with
%%   N$_\text{orient}=9$, (16, 16) pixels per cell, and (1, 1) cells
%%   per block.  The brightest lines align with prominent edges.
%%   Note, while fewer pixels per cell will better resolve the arc
%%   feature, the ``edge'' corresponding to the arc is less prominent.
%%   Bottom row: Visualized HOG feature vectors for the LSST-like
%%   single epoch and stacked images with the best scoring HOG
%%   parameterization of N$_\text{orient}=4$, (3, 3) pixels per cell,
%%   and (3, 3) cells per block.  The single epoch image (left) does
%%   not have significant enough edge features to emerge in the HOG,
%%   but the stacked image (right) does.}  \end{center} \end{figure*}
%%   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%-------------------------------------------------%
\subsubsection{Machine Learning Algorithm: Logistic Regression}\label{sec:LR}
%-------------------------------------------------%
The problem of detecting gravitational lenses in images falls under
the general category of {\em{classification}} in machine learning. In
general the task is to find a function that assigns data points $x$ to
one of two or more classes, denoted by the class label $y$.  This is
equivalent to specifying a decision boundary, or decision boundaries
between the classes in the space of the data points.  (Compare this to
{\em{regression}} in which the task is to find a function $y=f(x)$,
where $y$ is a continuous, rather than discrete, variable.) In our
case, we have two classes: lens and non-lens containing images, and
the data points $x$ are the HOG feature vectors transformed from the
training set data.  In this paper we use the Logisitic Regression (LR)
algorithm, for which the decision boundary is a hyperplane .  (The
equivalent in the regression setting would be linear regression.) In
LR we determine the optimal hyperplane by minimizing the objective
function
\begin{equation}
\label{eqn:log_reg_objective}
L(A,b) = \sum_i \log \left[1 + \exp \left(-y_i (A \cdot x_i + b) \right) \right]
\end{equation}
where $x_i$ is a data point (HOG feature vector), $y_i$ is the
known label for that data point ($1$ for a lens containing
image, $-1$ for a non-lens containing), and $A$ and $b$ are
the parameters of the hyperplane. Eq. \ref{eqn:log_reg_objective}
is to be minimized with respect to $A$ and $b$.
Other more complicated machine learning algorithms exist which
do not necessarily produce a linear decision boundary, such as
Support Vector Machines (SVM), Random Forests, and Neural Networks.
training set data \citep{hastie_09}.

>>>>

%-------------------------------------------------%
\begin{abstract} 
Gravitational lensing offers a direct probe of the underlying mass
distribution of lensing systems, a window to the high redshift
universe, and a geometric probe of cosmological models.  The advent of
large scale surveys such as the Large Synoptic Sky Telescope and
Euclid has prompted a need for automatic and efficient identification
of strong lensing systems.  We present (1) {\em (ALL) Automated
  Lensing Learner}, a strong lensing identification pipeline that will
be publicly released as open source software, and (2) a publicly
available mock LSST dataset with strong galaxy-galaxy lenses.  In this
first application of the pipeline, we employ a fast feature extraction
method, Histogram of Oriented Gradients (HOG), to capture edge
patterns that are characteristic of strong gravitational arcs in
galaxy-galaxy lensing.  We use logistic regression to train a
supervised classifier model on the HOG of HST- and LSST-like images.
We use the area under the curve (AUC) of a Receiver Operating
Characteristic (ROC) curve to assess model performance; $AUC=1.0$ is an
ideal classifier, and $AUC=0.5$ is no better than randomly guessing.
Our best performing models on a training
set of 10,000 lens containing images and 10,000 non-lens containing
images exhibit an AUC of 0.975 for an HST-like sample.  However, for
one orbit of LSST, our model only reaches an AUC of 0.625.  For
10-year mock LSST observations, the AUC improves to 0.809.  Model
performance appears to continually improve with the size of the
training set.  Models trained on fewer images perform better in
absence of the lens galaxy, the removal of which can decrease the
false positive rate due to the blending of light between source images
and the lens galaxy.  However, with larger training data sets,
information from the lens galaxy actually improves model performance.
Our results demonstrate an efficient and effective method for automatically
identifying strong lenses that captures much of the complexity of the
arc finding problem. The linear classifier both runs on a
personal laptop and can easily scale to large data sets on a computing
cluster, all while using existing open source tools.
\end{abstract}
%-------------------------------------------------%

@@

% Gravitational lensing as a geometric test of cosmology.  Also
% probes underlying matter distribution in galaxy clusters, sensitive
% to things like recent accretion and environment.
Gravitational lensing occurs when intermediate fluctuations in the
matter density field deflect light from background sources
\citep[see][for a review]{kneibandnatarajan_11}.  Strong gravitational
lensing can manifest as visible giant arcs magnifying high redshift
galaxies \citep{lyndsandpetrosian_86,gladders_etal03}, multiply imaged
quasars \citep{walsh_etal79}, multiply imaged galaxies
\citep{sharon_etal05}, and arclets \citep{bezecourt_etal98}.  Lensing
signatures probe the underlying dark matter distribution of the lens
\citep{warrenanddye_03}, and high redshift galaxy formation
\citep{allam_etal07}. They also provide a geometric test of cosmology via
comparison of predicted arc abundances with observed abundances
\citep{kochanek_96,chae_03,linder_04}, and via time-delay between signals
from multiply imaged quasars \citep{xli_etal12,suyu_etal14,suyu_etal16}.

% Application of constraining mass profiles - large numbers needed.
The application of strong gravitational lensing to constrain the mass
distribution of strong lenses, such as early-type galaxies (ETGs),
necessitates large samples of galaxy-galaxy lensing systems.
\citet{miraldaescudeandlehar_92} first suggested that massive
ellipticals would likely be frequent strong lensing sources in optical
surveys.  These systems contain a background source galaxy that the
lens galaxy deflects into a partial or full arc shaped Einstein ring.
The strong lensing signature directly probes the underlying matter.
The identification of such systems in upcoming surveys is the first
step in constraining the mass-to-light ratio for a large number of
objects in this mass range.

@@

% Transition to automated identification
Recent efforts have focused on the development of automated methods
with performance comparable to or better than humans. {\em SpaceWarps}
is a part of the Zooniverse Project \citep{marshall_etal16}, which
also includes {\em Galaxy Zoo}, the citizen science based image
classification of galaxy types \citep{lintott_etal08}.  Galaxy
classification is an early example in which machine
learning algorithms successfully trained models to classify astronomical
images with comparable performance to humans \citep{dieleman_etal15}.

% Early efforts
Earlier efforts on automated, or ``robot'', identification of strong
lensing systems have two distinct generalized steps.  The first
enhances and extracts characteristic features, and the second uses
some form of pattern recognition in the features to classify lens and
non-lens containing systems.  Among others, selected features might
include shape parameterization
\citep[see][e.g.]{,alard_06,kuboanddellantonio_08,xu_etal16,lee_17},
colors in multi-band imaging \citep{maturi_etal14,gavazzi_etal14},
light profiles \citep{braultandgavazzi_15}, and characteristics of
potential lens galaxies \citep{marshall_etal09}.  Pattern recognition
often incorporates cutoffs in selected parameter space distributions
\citep{lenzen_etal04,joseph_etal14,paraficz_etal16}.  A number of
these have been publicly distributed, with specific end applications.
For example, {\em ArcFinder} is one such code that finds arcs in
groups or cluster scale lens \citep{seidelandbartelmann_07}, and {\em
  RingFinder} is an analogous tool in searching for multiply imaged
quasars \citep{gavazzi_etal14}.  Codes like these have been
complementary to human identification \citep{more_etal16}.

% Classical machine learning methods
Pattern recognition methods have transitioned to using ``machine
learning'' algorithms in place of manual cutoffs.  With
machine learning, we can train a model to separate a dataset into
a known set of classes, such as ``lensing systems'' and
``non-lensing systems''. However, many classic machine learning
algorithms do not work well in image space, i.e. directly using the
raw pixel values of the image, but first require a process of
\em{feature engineering}, in which domain expertise is used to
extract variables which are more directly related to the classification
task at hand.  An example would be the application of an edge detector.
The optimal weights and cutoffs for these derived variables that are
used to determine the class label of an image are found automatically
by the algorithm.

% Recent works, neural nets, and deep learning
Some more recent works have made use of a subset of machine
learning algorithms called neural networks to classify images,
either from derived image
parameters or directly in image space.  In \citet{estrada_etal07},
authors used derived shape parameters to train neural nets to identify
arc candidates.  \citet{agnello_etal15} used neural networks trained
on data from multiband magnitudes, and \citet{bom_etal16} used
extracted morphological parameters.

% Convolutional and residual
The most advanced use of neural networks for strong lensing
classification operate directly in image space.
\citep{petrillo_etal17} used mock Kilo Degree Survey data to train
convolutional neural networks, with a training set size of six million
images. \citet{lanusse_etal17} used state-of-the-art deep residual
neural networks
to also work directly in the image space with minimal image
pre-processing.  A major strength of the \citet{lanusse_etal17}
implementation is that in comparison to deep convolutional neural
networks, deep residual neural networks have been found to
be easier to train and perform better.

%  Relevance of our work
While there has been a recent surge in the use of deep neural
networks applied to image classification problems in astronomy,
it is not always easy to scale these techniques to large data
sets, nor are the necessary computational resources and hardware,
such as GPUs, easily accessible to the
entire scientific community.  We present a supervised classification
pipeline built with open source tools, tested on mock HST and LSST
data, which will also be made publicly available.  Our goal is to
create a centralized open source pipeline that is general enough for
the user to select or add appropriate image processing techniques to
augment the features of the lensed image, train and test data with
different machine learning techniques, and to select the sequence of
methods that provides optimal model performance.  Results are
reproducible on personal computers, as both the pipeline and the
data will be publicly distributed.

% Summary of paper 
This first paper serves as both a presentation of the mock data set
and as a thorough case study for our pipeline with one feature
extraction and classical machine learning method, respectively, the
{\em histogram of oriented gradients} (HOG) and {\em logistic
  regression} (LR).  HOG is a fast feature extraction procedure that
quantifies edges in images, commonly used to identify humans in
security software.  LR is a linear classifier, making its scalability
relatively straightforward with existing open source tools.  This
framework is easily extendable to other feature extraction methods
and machine learning algorithms for supervised classification.

@@

There are 10,000 lens containing mock observations and 10,000
non-lens containing mock observations for running our parameter search. 
We keep a hold-out set of 1,000 lens containing mock observations
and 1,000 non-lens containing mock observations on which we evaluate
the final trained model.

@@

To produce simulated lensed galaxies, we first model the mass of lens
galaxies as a Singular Isothermal Ellipsoid (SIE). This model is
both analytically tractable and is consistent with models of
individual lenses and lens statistics on the length scales relevant
for strong lensing
\citep[e.g.]{koopmans_etal06,gavazzi_etal07,dye_etal08,liandchen_09}.

The normalized convergence map of the SIE model is defined as:
%-------------------------------------------------%
\begin{equation}\label{eqn:sie}
\kappa = \frac{\theta_{\rm E}}{2}\frac{1}{\sqrt{x_1^2/q+x_2^2 q}},
\end{equation}
%-------------------------------------------------%
where $\theta_{\rm E}$ is the Einstein Radius and $q$ is the axis ratio.
The Einstein radius can be calculated from the redshift of the
lens, the redshift of the source, and the velocity dispersion of the
lens galaxy as follows,
%-------------------------------------------------%
\begin{equation}
\theta_{\rm E} = 4\pi\left(\frac{\sigma_v}{c}\right)\frac{D_{ls}}{D_{s}}, 
\end{equation}
%-------------------------------------------------%
Here, $c$ is the speed of light, $\sigma_v$ is the velocity dispersion
of the lens galaxy, $D_{ls}$ and $D_{s}$ are respectively the angular
diameter distances from the source plane to the lens plane and from
the source plane to the observer.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure*}[t]
\begin{center}
\includegraphics[width=2\columnwidth]{figures/tiles/compilations/telescope_compilation.pdf}
\caption{Left to right show example mock HST, LSST 10 year, and LSST 1
  year images.  The top row corresponds to a lensing system with a
  very visible arc signature, and the bottom row to a lensing system
  that is less obvious.  Example mock HST images have
  $n_\text{pix}\times n_\text{pix}=300\times300$.  Example mock LSST
  images have $n_\text{pix}\times n_\text{pix}=45\times45$.  The
  resolution and noise of a ground based telescope is noticeably
  worse. Visual identification of giant arcs in the LSST images in the
  bottom row is very difficult.}\label{fig:mockimages}
\end{center}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

To perform our analysis, we have used tools from {\em Scikit-learn}
\citep{pedregosa_etal12}.  We outline the identification pipeline in
Figure~\ref{fig:pipeline}, which is a general description of
supervised classification.  Supervised classification is a class of
machine learning where the class labels in the training
set are known.  In our case, the labels are ``lens'' and ``non-lens''
containing images.

The first step of our pipeline consists of a feature extraction stage,
where our feature vector is a {\em histogram of oriented gradients}
(HOG) \citep{dalalandtriggs_05} that quantifies edges in the image.  We
describe the method and parameter search in Section~\ref{sec:hog}.  We
then use {\em Logistic Regression} (LR), a machine learning algorithm
described in Section~\ref{sec:LR}, to train a classifier model on a
subset of our images.  LR requires a parameter search over the
regression coefficient, $C_{LogReg}$, which we explore in
Section~\ref{sec:regularization}.  We briefly comment that our initial
tests with a {\em Support Vector Machine} (SVM) using radial basis
fucntions as an alternative
machine learning algorithm yielded negligible performance improvement,
and significantly increased computation cost.  This indicated that the
features of lens and non-lens images are relatively well separated by
hyperplanes in feature space.  For these reasons, we do not include
SVM in our final analysis and comparisons and continue all discussions
with a linear classifier.

@@

The second step of our analysis is to test our trained model on an
independent subset of the images to assess the model performance.
Here, we evaluate the model on each test image, predicting a
likelihood (``score'') between 0 and 1 that image contains a lensing
system.  This ``holdout set'' is not used in any of our parameter
searches to keep our test metric independent of tuning.

@@

%-------------------------------------------------%
\subsubsection{Feature Extractor: Histogram Oriented Gradients}\label{sec:hog}
%-------------------------------------------------%
% HOG intro - 
Originally created for human detection in computer vision, histogram of
oriented gradients (HOG) is a feature extraction method that computes
centered horizontal and vertical gradients. HOG is relatively robust
to noise in the image, and is a fairly fast transform that describes
edges.  Details can be found in \citet{dalalandtriggs_05}, but we
describe the procedure here.  The end result of HOG is a one
dimensional histogram computed as follows.

@@

% Initial binning
The orientation gradient of all pixels within each cell are binned
into the quantized orientations, providing a net gradient description
within that cell.  As an example, for $N_{orient}=3$, our bins are
centered at $\theta=0, 2\pi/3, 4\pi/3$ in radians.  If a cell
only has a gradient in the
$\theta=\pi/2$ direction, it will contribute 75\% of its magnitude to the
$\theta=2\pi/3$ bin, and 25\% of its magnitude to the $\theta=0$ bin.
The bins in all cells are then concatenated to make a larger feature
vector that is $N_{orient}$\times$N_{cells}$.

% Normalization
The last step is a normalization procedure to control for illumination
effects.  Here, the sub-histograms of each cell within the same block
are normalized with respect to one another before the transformation
returns the final feature vector.  The division of the image and the
quantization of orientations are thus controlled by 3 parameters in HOG:
$N_{orient}$, cells-per-block, and pixels-per-cell.  We discuss how we
select parameters using cross-validation in
Section~\ref{sec:gridsearch}

@@

Next, the cells per block parameter determines the normalization of
each cell with respect to the neighboring cell.  In general, this will
down-weight arc-like edges in cells that neighbor very bright cells,
such as cells that cover the central lens galaxy.  We therefore vary
the cells per block parameter between (2, 2) and (4, 4) for the
LSST-like images and between (1, 1) and (4, 4) for the HST-like
images.

@@

Finally, the resolution of the overall image will also limit the
additional information that an increase in $N_\text{orient}$ will
provide.  \todo{From the grid search, the best case number of orientations
for each dataset is
$N_\text{orient,HST}=9$, $N_\text{orient,LSST10}=3$, and
  $N_\text{orient,LSST1}=$.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure*}[t]
\begin{center}
\includegraphics[width=2\columnwidth]{figures/tiles/compilations/hog_telescope_compilation.pdf}
\caption{Left to right show example image transforms of mock images
  from Figure~\ref{fig:mockimages} with a visualized histogram of
  oriented gradients.  The image transform picks up edge features,
  with arc features showing up as edges across radial orientations.
  \wording{Each of the oriented gradients within a cell is
    color-coded by magnitude, and represented as a line in the
    direction perpendicular to that gradient.}  The actual extracted
  features fill a one-dimensional feature vector comprised of the
  magnitudes of each of the oriented edges within the visualized
  cells.}\label{fig:hogimages}
\end{center}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%            
%% \caption{Top row: Visualized HOG feature vectors of the sample
%%   lensed HST-like image from Figure~\ref{fig:mockimages}.  Left:
%%   The corresponding HOG visualization with N$_\text{orient}=5$,
%%   (32, 32) pixels per cell, and (1, 1) cells per block.  Right: The
%%   HOG visualization of the same mock image with
%%   N$_\text{orient}=9$, (16, 16) pixels per cell, and (1, 1) cells
%%   per block.  The brightest lines align with prominent edges.
%%   Note, while fewer pixels per cell will better resolve the arc
%%   feature, the ``edge'' corresponding to the arc is less prominent.
%%   Bottom row: Visualized HOG feature vectors for the LSST-like
%%   single epoch and stacked images with the best scoring HOG
%%   parameterization of N$_\text{orient}=4$, (3, 3) pixels per cell,
%%   and (3, 3) cells per block.  The single epoch image (left) does
%%   not have significant enough edge features to emerge in the HOG,
%%   but the stacked image (right) does.}  \end{center} \end{figure*}
%%   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%-------------------------------------------------%
\subsubsection{Machine Learning Algorithm: Logistic Regression}\label{sec:LR}
%-------------------------------------------------%
The problem of detecting gravitational lenses in images falls under
the general category of {\em{classification}} in machine learning. In
general, the task is to find a function that assigns data points $x$ to
one of two or more classes, denoted by the class label $y$.  This is
equivalent to specifying a decision boundary, or decision boundaries
between the classes in the space of the data points.  (Compare this to
{\em{regression}} in which the task is to find a function $y=f(x)$,
where $y$ is a continuous, rather than discrete, variable.) In our
case, we have two classes: lens and non-lens containing images, and
the data points $x$ are the HOG feature vectors extracted from the
images.  In this paper we use the \em{Logisitic Regression}
(LR) algorithm, for which the decision boundary is a hyperplane.  (The
equivalent in the regression setting would be linear regression.) In
LR we determine the optimal hyperplane by minimizing the objective
function
\begin{equation}
\label{eqn:log_reg_objective}
L(A,b) = \sum_i \log \left[1 + \exp \left(-y_i (A \cdot x_i + b) \right) \right]
\end{equation}
where $x_i$ is a data point (HOG feature vector), $y_i$ is the
known label for that data point ($1$ for a lens containing
image, $-1$ for a non-lens containing), and $A$ and $b$ are
the parameters of the hyperplane. Eq. \ref{eqn:log_reg_objective}
is to be minimized with respect to $A$ and $b$.
Other more complicated machine learning algorithms exist which
do not necessarily produce a linear decision boundary, such as
\em{Support Vector Machines} (SVM), \em{Random Forests}, and
\em{Neural Networks} \citep{hastie_09}.

The HOG feature vectors in this paper can be very high-dimensional.
When dealing with high-dimensional data, where the number of
dimensions becomes comparable to the number of data points,
overfitting can become an issue. (An example of an extreme case
of overfitting is fitting a degree $n$ polynomial to $n$
points.  The polynomial would simply wiggle so that it goes
through every point, and would have no predictive power if you
tried to interpolate or extrapolate.) In machine learning,
overfitting is avoided using different {\em{regularization}}
techniques. A common choice for logistic regression is to
add a penalty term to Equation~\ref{eqn:log_reg_objective}
\begin{equation}\label{eqn:log_reg_regularized}
L_\text{Reg}(A,b,C_\text{LogReg}) = L(A,b) + \frac{1}{2C_\text{LogReg}} \|A\|
\end{equation}
where the norm $\|A\|$ is typically taken as either the
sum of squares of the coefficients ($L_2$ norm) or the sum
of absolute values of the coefficients ($L_1$ norm).  In this
paper we use the former.
